Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2023-12-23 alignpair_letter.tex
Add new COATi FST figure.

Statistical alignment is typically performed using pairwise hidden Markov models (pair-HMMs), which have the ability to rigorously model molecular sequence evolution \citep{bradley2007transducers}.
Pair-HMMs are computational machines with two output tapes. Pair-HMMs contain a finite number of states---typically labeled match, insert, and delete---that emit symbols (nucleotides or amino acids) to one or both tapes.
Each tape represents a sequence, and a path through a pair-HMM is a possible pairwise alignment.
Conceptually, these machines generate two sequences ($\text{X}$ and $\text{Y}$) from an unknown ancestor and can calculate the probability that two sequences are related, represented by $P(\text{X}, \text{Y})$ \citep{yoon_2009_hmm}.

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=1]{figures/fig-fst-base-calling-new.pdf}
    \caption[DNA Transcription via FST Composition]{DNA transcription via FST composition. The composition of (a) a codon complimenting FST and (b) an FST that replaces T with U generates (c) an FST that implements codon transcription ($a~\circ~b = c$).}
    \label{fig:transcription}
\end{figure}

>>>>

In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{Lunter2005-hk}. While approaches vary, a statistical aligner for a pair of sequences $X$ and $Y$ typically finds an alignment $A$ that maximizes the joint probability $P(X, Y, A)$ or samples alignments from the posterior $P(A | X, Y)$. This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{bradley2007transducers}. Pair-HMMs are computational machines with two output tapes. Each tape represents a sequence, and a path through a pair-HMM is a possible pairwise alignment.
Conceptually, these machines generate two sequences from an unknown ancestor and can calculate the probability that two sequences are related, i.e.\ $P(X, Y)$ \citep{yoon_2009_hmm}.

While the use of pair-HMMs is ubiquitous in bioinformatics, their

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{figures/fig-fst-base-calling-new.pdf}
    \caption[FSTs more the generation of sequences]{%
    Finite state transducers (FSTs) model the generation of an output sequences based on an input sequence.
    %
    (a) A graph of a probabilistic FST \citep{cotterell-etal-2014-stochastic} for base-calling errors using a Mealy-machine architecture, where parameter $u$ is the error rate. This graph contains two states (S and M) connected by arcs, with labels ``input symbols~:~output symbols~/~weight''. Arcs consume symbols from the input sequence and emit symbols to the output sequence. Weights describe the probability that an arc is taken given the input symbols. Epsilon (ε) is a special symbol denoting that no symbols were either consumed or emitted.
    %
    (b) An FST for matching sequences against ambiguous nucleotides (N). This FST is not a true probabilistic FST and cannot be used to simulate output sequences since it is missing a parameter to control how often Ns are added to the output sequence.
    %
    (c) An FST that results from the composition ($∘$ operation) of the Error FST with the Ambiguity FST. As with (b), this composed FST cannot be used to simulate output sequences; however, it does properly weight the ambiguous nucleotide N as representing any other symbol.
    %
    %Let $x$ be a symbol in the input of (a), $y$ be a symbol in the output of (b), and $z$ be a symbol in both the output of (a) and the input of (b). An arc in the composed FST (c) has an input symbol $x$, an output symbol $y$, and a weight that is the combination over $z$ of the aggregate weights of all $x$-$z$-$y$ arc pairs.
    %
    }\label{fig:base-calling}
\end{figure}

### 2023-12-23 figures/fig-fst-base-calling-new.tex
Add new COATi FST figure.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
\usepackage{fontspec}
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{xcolor}
\usepackage{xfrac}

\usetikzlibrary{shapes,arrows,positioning,calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{decorations.text}
\usetikzlibrary{hobby,spath3,intersections}

@@

\tikzset{
  end/.style={fill=none, text=black, inner sep=1pt},
  block/.style={circle,on grid,fill=colorBu,text=white,minimum width=2em,node distance=4em},
  line/.style={draw, -stealth', thick},
  lab/.style={text = black, inner sep=1pt},
  title/.style={text = black, font=\bfseries}
}

@@

\node[block,right = of start] (s) {S}; 
\node[block,left = of end] (m) {M};

\draw[line] (start) to (s);
\draw[line] (m) to (end);

@@

\foreach \y [count=\yi] in {
  {A:A/$(1-e)$}, {A:C/$(e/3)$}, {A:G/$(e/3)$}, {A:T/$(e/3)$},
  {C:A/$(e/3)$}, {C:C/$(1-e)$}, {C:G/$(e/3)$}, {C:T/$(e/3)$}
} {
    \coordinate (y) at ($(mid) + (0, 15.5em) + \yi*(0, -1.75em)$);
    % \path[spath/save=arc] (s.center) to[curve through={(y)}] (m.center);

@@

\foreach \y [count=\yi] in {
  {G:A/$(e/3)$}, {G:C/$(e/3)$}, {G:G/$(1-e)$}, {G:T/$(e/3)$},
  {T:A/$(e/3)$}, {T:C/$(e/3)$}, {T:G/$(e/3)$}, {T:T/$(1-e)$}
} {
    \coordinate (y) at ($(mid) + (0, 0.25em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=north] at (y) {\y};
}

@@

\foreach \y [count=\yi] in {
  {A:N/$1$}, {A:A/$(1-e)$}, {A:C/$(e/3)$}, {A:G/$(e/3)$}, {A:T/$(e/3)$},
  {C:N/$1$}, {C:A/$(e/3)$}, {C:C/$(1-e)$}, {C:G/$(e/3)$}, {C:T/$(e/3)$}
} {
    \coordinate (y) at ($(mid) + (0, 19em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=south] at (y) {\y};
}

\foreach \y [count=\yi] in {
  {G:N/$1$}, {G:A/$(e/3)$}, {G:C/$(e/3)$}, {G:G/$(1-e)$}, {G:T/$(e/3)$},
  {T:N/$1$}, {T:A/$(e/3)$}, {T:C/$(e/3)$}, {T:G/$(e/3)$}, {T:T/$(1-e)$}
} {
    \coordinate (y) at ($(mid) + (0, 0.25em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=north] at (y) {\y};
}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
\usepackage{fontspec}
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{xcolor}

\usetikzlibrary{shapes,arrows.meta,positioning,calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{decorations.text}
\usetikzlibrary{hobby,spath3,intersections}

@@

\tikzset{
  end/.style={fill=none, text=black, inner sep=1pt},
  block/.style={circle,on grid,fill=colorBu,text=white,minimum width=2em,node distance=4em},
  line/.style={draw, -Stealth, thick},
  lab/.style={text = black, inner sep=1pt},
  title/.style={text = black, font=\bfseries}
}

@@

\node[block,right = of start] (m) {M}; 
\node[block,left = of end] (s) {S};

\draw[line] (start) to (m);
\draw[line] (s) to (end);

@@

\foreach \y [count=\yi] in {
  {A:A/$(1-u)$}, {A:C/$(u/3)$}, {A:G/$(u/3)$}, {A:T/$(u/3)$},
  {C:A/$(u/3)$}, {C:C/$(1-u)$}, {C:G/$(u/3)$}, {C:T/$(u/3)$}
} {
    \coordinate (y) at ($(mid) + (0, 15.5em) + \yi*(0, -1.75em)$);
    % \path[spath/save=arc] (s.center) to[curve through={(y)}] (m.center);

@@

\foreach \y [count=\yi] in {
  {G:A/$(u/3)$}, {G:C/$(u/3)$}, {G:G/$(1-u)$}, {G:T/$(u/3)$},
  {T:A/$(u/3)$}, {T:C/$(u/3)$}, {T:G/$(u/3)$}, {T:T/$(1-u)$}
} {
    \coordinate (y) at ($(mid) + (0, 0.25em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=north] at (y) {\y};
}

@@

\foreach \y [count=\yi] in {
  {A:N/$1$}, {A:A/$(1-u)$}, {A:C/$(u/3)$}, {A:G/$(u/3)$}, {A:T/$(u/3)$},
  {C:N/$1$}, {C:A/$(u/3)$}, {C:C/$(1-u)$}, {C:G/$(u/3)$}, {C:T/$(u/3)$}
} {
    \coordinate (y) at ($(mid) + (0, 19em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=south] at (y) {\y};
}

\foreach \y [count=\yi] in {
  {G:N/$1$}, {G:A/$(u/3)$}, {G:C/$(u/3)$}, {G:G/$(1-u)$}, {G:T/$(u/3)$},
  {T:N/$1$}, {T:A/$(u/3)$}, {T:C/$(u/3)$}, {T:G/$(u/3)$}, {T:T/$(1-u)$}
} {
    \coordinate (y) at ($(mid) + (0, 0.25em) + \yi*(0, -1.75em)$);
    \path[line,shorten <=1em+0.2pt, shorten >=1em+0.3pt] (s.center) to[curve through={(y)}] (m.center);
    \node[lab,anchor=north] at (y) {\y};
}
